A comparison of different methods for calculating tangent-stiffness matrices in a massively parallel computational peridynamics code.

### 2013-10-27 jacobian_calc_CMAME.tex
Through with Peridigm section

In order to maintain the quadratic convergence properties of the first-order Newton's method in quasi-static nonlinear analysis of solid structures it is crucial to obtain accurate, algorithmically consistent tangent-stiffness matrices. For an extremely small class of nonlinear material models, these consistent tangent-stiffness operators can be derived analytically; however, most often in practice, they are found through numerical approximation of derivatives.

@@

This section contains background information on the differentiation techniques underlying the tangent-stiffness matrix calculation methods compared in the study. Since first-order finite difference techniques are considered well known, they are not described here. Instead the reader is referred to \cite[Chap. 4.1.3]{chapra2010}.

@@

\subsubsection{Differentiation Technique: Automatic-differentiation} 
\label{ADsubsection}

@@

When using automatic-differentiation to calculate a tangent-stiffness matrix, one only needs to follow the definition of the tangent-stiffness matrix and issue the correct commands to the AD system. The formula is just as expected:

\begin{equation} K_{ij} = \frac{\partial F_i^{int}(\vec{u})}{\partial u_j} \end{equation}

\subsubsection{Solve a Nonlinear System with a Tangent-stiffness Matrix} \label{solve_a_system}
The reader is referred to a succinct explanation of the Newton Raphson method, which shows the
role of a tangent-stiffness matrix or Jacobian in that solution method \cite[chap.
13]{young2009}.

\section{Methods and Materials}

\subsection{Material: Perdynamics Code}

The following subsubsections identify the program modified to run the simulations for the
comparative study and explain what modifications were necessary.

\subsubsection{\emph{Perdigm} and Peridynamics} 
\subsubsection{Implementing Complex-Step in \emph{Peridigm}}

Complex step was implemented in \emph{Perdigm} as follows. First a way of
persistently allocating memory needed to store imaginary values for nodal
position and dilatation was determined. Persistent memory allocation is a very
basic way to improve performance in general for a computer program. The way
memory was allocated for similar vector and scalar variables used by the prior
existing forward difference method was copied and new 'field spec' variables
were added into the 'elastic' material model to store imaginary position
coordinates and imaginary dilatation. Because of \emph{Peridigm}'s dependence
on the \emph{Epetra} package of the \emph{Trilinos} library, 'field spec'
variables could not be simply be declared as having a complex datatype. Program
methods that were used in the course of applying the forward-difference method
were copied, renamed and re-written to follow the CTSE formula, which involved
changing the datatype of intermediate variables which were locally scoped and
dynamically casting persistent variables, like the 'field spec' variables, to
complex datatypes where necessary. Additional modifications were for the
purpose of collecting experimental data and are not part of complex-step
itself. The source code for implementing complex-step in \emph{Perdigim} can be
found on the corresponding author's website.  Because \emph{Peridigm} was
written by computer scientists rather than mechanical engineers primarily, many
advanced techniques in C++ such as multiple virtual inheritance, pointer
arithmetic and distributed data structures were used in the basic code and by
necessity in the modified code.  To see complex-step in a more narrow context,
it is advised that readers also take a look at the examples of complex-step and
the other methods provided on the corresponding author's website.

@@

\begin{figure}[h]
  \centering
  \scalebox{1.0}{\input{./figs/serial_speed.pgf}}
  \caption{Serial test series speed measurements.}
  \label{fig:serial_speed}
\end{figure}

@@

\begin{figure}[h]
  \centering
  \scalebox{1.0}{\input{./figs/multi_speed.pgf}}
  \caption{Multicore test series speed measurements.}
  \label{fig:multi_speed}
\end{figure}

@@

\begin{figure}[h]
  \centering
  \scalebox{1.0}{\input{./figs/serial_accuracy.pgf}}
  \caption{Serial test series accuracy measurements.}
  \label{fig:serial_accuracy}
\end{figure}

@@

\begin{figure}[h]
  \centering
  \scalebox{1.0}{\input{./figs/multi_accuracy.pgf}}
  \caption{Multicore test series accuracy measurements.}
  \label{fig:multi_accuracy}
\end{figure}

@@

\begin{figure}[h]
  \centering
  \scalebox{1.0}{\input{./figs/serial_speed_rel.pgf}}
  \caption{Serial test series efficiency measurements.}
  \label{fig:serial_efficiency}
\end{figure}

@@

\begin{figure}[h]
  \centering
  \scalebox{1.0}{\input{./figs/multi_speed_rel.pgf}}
  \caption{Multicore test series efficiency measurements.}
  \label{fig:multi_efficiency}
\end{figure}

>>>>

In order to maintain the quadratic convergence properties of the first-order Newton's method \cite{belytschko1999nonlinear} \cite[Ch.~13]{young2009} in quasi-static nonlinear analysis of solid structures it is crucial to obtain accurate, algorithmically consistent tangent-stiffness matrices. For an extremely small class of nonlinear material models, these consistent tangent-stiffness operators can be derived analytically; however, most often in practice, they are found through numerical approximation of derivatives.

@@

This section contains background information on the differentiation techniques underlying the tangent-stiffness matrix calculation methods compared in the study. Since first-order finite-difference techniques are considered well known, they are not described here. Instead the reader is referred to \cite[Chap. 4.1.3]{chapra2010}.

@@

\subsubsection{Automatic-differentiation} 
\label{ADsubsection}

@@

When using automatic-differentiation to calculate a tangent-stiffness matrix, one only needs to follow the definition of the tangent-stiffness matrix and issue the correct commands to the AD system. The formula is the same as the continuum formula
%
\begin{equation} 
  K_{ij} = \frac{\partial F_i^{int}(\vec{u})}{\partial u_j},
\end{equation}
%
but, with the caveat that $F^{int}$ is evaluated in a way that is algorithmically consistent, and therefore should yield quadratic convergence with Newton's method.

\section{Description of analysis tools and approach.} 
%
\subsection{Peridigm and  peridynamics.}
%
Comparisons of the different tangent-stiffness calculation techniques were carried out through code-development and \emph{in situ} instrumentation of the computational peridynamics code \emph{Peridigm} \cite{peridigm}. \emph{Peridigm} is distributed by Sandia National Laboratories as its primary open-source computational peridynamics code. It is a massively-parallel simulation code for implicit and explicit multi-physics simulations primarily used for solid mechanics and material failure. \emph{Peridigm} is a C++ code utilizing agile software components from Sandia's  \emph{Trilinos} project \cite{trilinos}. It's important to note that there is no specialization of the tangent-stiffness calculation methods described previously to this particular code or more generally to a computational peridynamics approach and the analysis carried out in this study should provide insight into the speed and accruacy of these methods when implemented into other computational mechanics analysis tools as well.  \emph{Peridigm} was chosen because of the author's relationship with the code development project and primarily because the FD, CD, and AD methods for tangent-stiffness calculation where areleady implemented in the code, leaving only the CS method to for development. However, because peridynamics is a nonlocal theory, the tangent-stiffness matrices have a much higher bandwidth (i.e. less sparsity) than traditional finite-element tangent-stiffness computations and the time required to construct the tangent-stiffness is a majority of the total computation time in any quasi-static or implicit dynamics analysis; therefore, the \emph{Peridigm} development team has an interest in profiling the performance of the tangent-stiffness computation methods.

Briefly, peridynamics \cite{silling2000ret, silling:psa, silling2010peridynamic} is a nonlocal reformulation of the partial-differential equations that provide the statement of momentum balance in classical continuum mechanics. Its primary goal is the avoid the use of spatial derivatives in the balance equations or constitutive laws such that discontinuous displacements, i.e. cracks, are mathematically consistent with the governing equations. It has demonstrated great promise in modeling problems with massive pervasive failure \cite{littlewood2010}, interesting phenomenon such as dynamic crack branching \cite{ha2010sod}, and is being extended to multiphyics phenomenon \cite{bobaru2011peridynamic,katiyar2013} and coupling to molecular simulations \cite{seleson2009peridynamics}.

\subsubsection{Implementing the complex-step method in Peridigm} 
%
The CS method was implemented in \emph{Perdigm} in a manner that follows closely the implementation of the FD technique.  However, because of \emph{Peridigm}'s reliance on \emph{Trilinos} and specifically the \emph{Epetra} package, special steps had to be followed to allow the use of complex number data types.  This is because \emph{Epetra} vectors, which can be thought of as distributed memory parallel cousins to standard C++ STL vectors, are hard coded to be of type {\tt double} and can not be simply be declared as having a complex datatype. Therefore, program methods that were used in the course of applying the FD method were copied, renamed and re-written to follow the CS formula, which involved changing the datatype of intermediate variables which were locally scoped and dynamically casting persistent memory variables to complex datatypes where necessary. We made no atttempt to ``tune'' the CS code in any way for performance, we attempted to implement the algorithm is such as way that it was in one-to-one correspondence with the FD and CD methods; however, it must be noted that the overhead associated with the dynamic casts in the CS method likely hurt its overall performance and this issue could be alleviated by using the next generation templated \emph{TPetra} vectors available in \emph{Trilinos} that are capable of taking an explicit instantiation of any datatype including complex.  There where additional modifications for the purpose instrumenting the code to collect comparitive data and are not specifically part of CS or other methods. The source code patch to \emph{Peridigm} for implementing CS can be found on the corresponding author's website.

\emph{Peridigm} was primarily written by expert computational scientists; therefore, many advanced techniques in C++ such as multiple virtual inheritance, pointer arithmetic, and distributed data structures were used in the basic code and by necessity in the modified code.  To see CS and other methods in a more narrow context, it is advised that readers also take a look at the simple examples provided on the corresponding author's website.

@@

\begin{figure}[tbp]
  \centering
  \scalebox{1.0}{\input{./figs/serial_speed.pgf}}
  \caption{Serial test series speed measurements.}
  \label{fig:serial_speed}
\end{figure}

@@

\begin{figure}[tbp]
  \centering
  \scalebox{1.0}{\input{./figs/multi_speed.pgf}}
  \caption{Multicore test series speed measurements.}
  \label{fig:multi_speed}
\end{figure}

@@

\begin{figure}[tbp]
  \centering
  \scalebox{1.0}{\input{./figs/serial_accuracy.pgf}}
  \caption{Serial test series accuracy measurements.}
  \label{fig:serial_accuracy}
\end{figure}

@@

\begin{figure}[tbp]
  \centering
  \scalebox{1.0}{\input{./figs/multi_accuracy.pgf}}
  \caption{Multicore test series accuracy measurements.}
  \label{fig:multi_accuracy}
\end{figure}

@@

\begin{figure}[tbp]
  \centering
  \scalebox{1.0}{\input{./figs/serial_speed_rel.pgf}}
  \caption{Serial test series efficiency measurements.}
  \label{fig:serial_efficiency}
\end{figure}

@@

\begin{figure}[tbp]
  \centering
  \scalebox{1.0}{\input{./figs/multi_speed_rel.pgf}}
  \caption{Multicore test series efficiency measurements.}
  \label{fig:multi_efficiency}
\end{figure}
